On using the decision trees to identify the local extrema in parallel global optimization algorithms

### 2023-02-01 2023/PAVT/ParallelDecisionTree.tex

\begin{document}
\mainmatter              % start of a contribution
%
\title{On the use of decision trees to identify the local extrema in parallel global optimization algorithms}
%
\titlerunning{Global Optimization ...}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{K.~A.~Barkalov\inst{1}\letter\orcidID{0000-0001-5273-2471} \and I.~G.~Lebedev\inst{1}\letter\orcidID{0000-0002-8736-0652} \and D.~I.~Silenko\inst{1}\letter\orcidID{0000-0002-2578-9699}}
%
\authorrunning{K.~A.~Barkalov et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{K.~A.~Barkalov and I.~G.~Lebedev, and D.~I.~Silenko}
%
\institute{Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, Russia \\
\email{ \{konstantin.barkalov,ilya.lebedev,silenko\}@itmm.unn.ru}
}

@@

@@

With regard to the search in a direction, it consists in performing a step from the base point found  during the investigating search along the straight line direction connecting this one with the previous  base point. The magnitude of this step is defined by a parameter set in advance.

\begin{figure}[!h] 
	\begin{center} 
		\begin{minipage}[h]{0.8\linewidth} 
			\includegraphics[width=1\linewidth]{figure/fig1.pdf} 
			\caption{An example of iterations of the Hooke-Jeeves algorithm} %%  подпись к рисунку 
			\label{fig:fig1} 
		\end{minipage} 
	\end{center} 
\end{figure}

 Fig. (\ref{fig:fig1}) presents small example of how the algorithm works. Function contour plot are displayed, filled circles indicate successful steps, unfilled circles indicate unsuccessful steps in the investigating search.

@@

Below, the results of comparing two parallel algorithms --- asynchronous global search algorithm  (AGSA) and its modification using the decision tree to find the attractor areas of the local minima  (Decision Tree) are presented. The numerical comparison was performed on two classes of the GKLS  functions (Simple and Hard from \cite{Sergeyev2006}) of dimensionalities 2, 3, 4, and 5. The stop  criterion for the algorithms was the falling of the next trial points into the $\varepsilon$-nearness of  true global minimum.  In Table \ref{tab:2}, the averaged numbers of iterations executed by the algorithms when solving the  problems and the values of time speedup relative to the sequential run are compared. The parallelization was performed using MPI technology, the run was performed on 8 processes.

@@

As one can see from Table \ref{tab:2},  the iteration speedup is large enough for the problems of any  dimensionality.

So far, as a result of the work done, we succeeded to combine the global search algorithm with a local  optimization method.  Unlike the known multistart schemes, the decision to run the local method is made using the decision  tree. The use of such a combination of methods allows considerably accelerating the algorithm.

@@

>>>>

\begin{document}
\mainmatter              % start of a contribution
%
\title{On using the decision trees to identify the local extrema in parallel global optimization algorithms}
%
\titlerunning{On using the decision trees to identify the local extrema }  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Konstantin Barkalov\inst{1}\letter\orcidID{0000-0001-5273-2471} \and Ilya Lebedev\inst{1}\letter\orcidID{0000-0002-8736-0652} \and Dmitry Silenko\inst{1}\letter\orcidID{0000-0002-2578-9699}}
%
\authorrunning{Konstantin Barkalov et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{K.A.~Barkalov and I.G.~Lebedev, and D.I.~Silenko}
%
\institute{Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, Russia \\
\email{ \{konstantin.barkalov,ilya.lebedev,silenko\}@itmm.unn.ru}
}

@@

\begin{figure} 
	\begin{center} 
		\begin{minipage}[h]{0.8\linewidth} 
			\includegraphics[width=1\linewidth]{figure/fig1.pdf} 
			\caption{An example of iterations of the Hooke-Jeeves algorithm} %%  подпись к рисунку 
			\label{fig:fig1} 
		\end{minipage} 
	\end{center} 
\end{figure}

@@

Fig. \ref{fig:fig1} shows the example of the algorithm work. Function level lines are displayed, filled circles indicate successful steps, unfilled circles correspond to unsuccessful steps performed during the investigating search. With regard to the search in a selected direction, it consists in performing a step from the base point found during the investigating search along the line connecting this one with the previous base point. The magnitude of this step is defined by a parameter set in advance.

@@

Below, the results of comparing two parallel algorithms --- asynchronous global search algorithm  (AGSA) and its modification using the decision tree to find the attractor areas of the local minima  (Decision Tree) are presented. The numerical comparison was performed on two classes of the GKLS  functions (Simple and Hard from \cite{Sergeyev2006}) of dimensionalities 2, 3, 4, and 5. 
These two classes differ in the size of the attraction region of the global minimum point. 
For the simple class of problems, the radius of the attraction region is three times larger than for a complex one.

The stop  criterion for the algorithms was the falling of the next trial points into the $\varepsilon$-nearness of  true global minimum.  In Table \ref{tab:2}, the averaged numbers of iterations executed by the algorithms when solving the  problems and the values of time speedup relative to the sequential run are compared. The parallelization was performed using MPI technology, the run was performed on 8 processes.

@@

As one can see from Table \ref{tab:2},  the iteration speedup is large enough for the problems of any  dimensionality. 
So far, as a result of the work done, we succeeded to combine the global search algorithm with a local  optimization method.  Unlike the known multistart schemes, the decision to run the local method is made using the decision  tree. The use of such a combination of methods allows considerably accelerating the algorithm.

@@

In addition to using decision trees to identify attraction regions for local extrema of multiextremal functions, we also plan to use machine learning methods to separate the variables of the problem being solved. In many applied optimization problems the dependence of the objective function on some parameters is either linear or unimodal. Separating such variables into a special group and solving the problem using a parallel recursive optimization scheme \cite{Barkalov2020_1} can reduce the time for solving the problem by orders of magnitude compared to using a global search for all variables at once.

\medskip

\textbf{Acknowledgments}. This study was supported by the Russian Science Foundation, project No.\,21-11-00204 .
